Personalised Map Matched Imputation: an approach to impute missing data in smartphone location logs

### 2018-04-29 manuscript/apaManu.Rmd
APA manuscript

---
title: "Manuscript"
author: "Boaz Sobrado"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(ggthemes)
library(ggplot2)
library(knitr)
```

# Abstract
*Personal mobility, or how people move in their environment, is associated with a vast range of behavioural traits and outcomes, such as socioeconomic status, personality and mental health. The widespread adoption of location-sensor equipped smartphones has generated a wealth of objective personal mobility data. Nonetheless, smartphone collected personal mobility data has remained underutilised in behavioural research, partly due to the practical difficulties associated with obtaining the data and partly because of the methodological complexity associated with analysing it. Recent changes in European regulation have made it easier for researchers to obtain this data, but the methodological difficulties remain. The difficulty lies in that smartphone location data is irregularly sampled, sparse and often inaccurate.  This results in a high proportion of missing data and significant noise. In this paper we present a method called Personal Map Matched Imputation (PPMI) to deal with missing data and noise in smartphone location logs. The main innovation of PPMI is that it creates a personalised spatial map for each individual based on all the available data. In doing so PPMI leverages the regularity of human mobility in order to smoothen noisy measurments and impute missing data values. By simulating missing periods in real data we find that a simple implementation of PPMI performs as well as existing methods for short (5 minute) missing intervals and substantially better for longer (1 day) missing intervals. When imputing a subset of real  missing data where travel logs are available as a reference points, we find that PPMI performs substantially better than existing models.*

@@

Why is human mobility important? How people move about in their environment affects a wide range of outcomes, such as health, income and social capital [@goodchild_toward_2010]. Therefore it is unsurprising that social scientists in numerous fields and even policymakers are interested in human mobility measures. The most widely administered personality questionnaires ask individuals to what extent they agree with statements such as "I love large parties", "I prefer going to the movies to watching videos at home" and "I love to travel to places that I have never been before" [@ipip]. In economic research, the postal code of an individual's home address is often used as a proxy for socioeconomic status [e.g. @villanueva]. Moreover, economists study geographic labour mobility extensively [e.g. @tatsiramos]. In fact, the European Commission considers labour mobility important enough to warrant its own directorate [@europeanComission]. The extensive use of measures like these across different domains strongly suggests that mobility metrics are linked to real world outcomes. Perhaps it is unsurprising that behavioural researchers have found that mobility measures can be used to predict academic performance [@wang_smartgpa:_2015], the incidence of obesity [@zenk_how_2009] or even the onset of a depressive epsiode in bipolar depression patients [@palmius_detecting_2017]. Indeed, there is an argument to be made that perhaps psychologists have not been studying mobility enough. When studying behavioural differences within individuals behavioural scientists have often neglected the fact that individuals vary not only over time but also over space. To fully understand behaviour we must understand how behaviour can vary accross environments.

Despite the importance of mobility measures, the majority of these measures are obtained through the use of questionnaires (such as the IPIP) or specifically through pen-and-paper travel diaries. Questionnaires and travel diaries are have well-known methodological flaws: they are burdensome to collect and rely on accurate self-reporting. The burdensomeness stems from the fact that participans must be explicitly asked to report on their movement patterns at frequent intervals due to forgetfulness. This makes the data expensive to collect and limits the extent of data collection in practice. In addition, the frequent reporting duties of the participant may bias the participants behaviour. Participant forgetfulness also limits the accuracy of self reported measures. There is clear evidence that participants are systematically biased when self-reporting mobility measures. For instance, participants underreport the frequency of short trips [@wolf_impact_2003] and underestimate the duration of regular commutes [@delclos-alio_keeping_2017]. These obstacles can be overcome by using objective data on human mobility.

Social scientists now have the unprecedented opportunity to easily obtain objective data on human mobility from smartphones. Before smartphones the only way to collect objective data on human mobility involved giving participants an expensive professional-grade location sensor and convincing them to take it with themselves at all times. @barnett_inferring_2016 points out that introducing a new device to the participant's life may bias their behaviour. Moreover, collecting data in such a way is costly, places a high burden on participants and therefore the logs do not span a long time [@barnett_inferring_2016. Today millions of individuals carry smartphones with themselves every day and do not need to be encouraged to do so by researchers. Smartphones are equipped with a range of sensors that can be used to track the location of the device at all times. These smartphones can collect and store hundreds of location measurements a day. For instance,  Google Location History contains movement information on millions of users, often spanning years [@location_history_timeline_2017]. Moreover, recent changes in EU regulations with regard to consumer data-portability rights ensure that a willing participant should be able to easily share this information with researchers at no cost to the either the participant or the researchers [@commission_protecting_2017]. Taken together, this means that researchers now have at their disposal the ability to easily access the objective human mobility data of millians of individuals spanning several years at little cost and without a significant burden of the participants.

@@

Smartphone location measures are obtained primarily (but not exclusively) by Global Positioning System (GPS) measurements. A GPS sensor uses the distance between a device and several satellites to determine the location of the device. Although using a GPS sensors is the most accurate way to establish location on a smartphone, the GPS sensor is the most energy consuming sensor on most smartphones [@lamarca_place_2005; @chen_practical_2006]. In order to avoid battery depletion and to overcome computational constraints smartphones also use less-accurate heuristics such as WiFi access points and cellphone tower triangulation. Smartphone location logs contain measurements from all of these sources, usually in the form of timestamped lattitude, longitude and accuracy values. The accuracy $a$ of any given measurement is given in meters such that it represents the radius of a 67% confidence circle [@location_history_timeline_2017]. In other words, the true location of a device should be within the radius $a$ of the measurement 67% of the time.

@@

In order to work with secondary logs, researchers need to be able to handle the data sparcity that leads to missing data. Missing data is a pervasive issue in secondary logs as it can arise due to several reasons. Technical reasons include signal loss, battery failure and device failure. Behavioural reasons include leaving the phone at home or switching the device off. As a result, secondary logs often contain wide temporal gaps with no measurements. For instance, several research groups studying mental health report missing data rates between 30% to 50% [@saeb_mobile_2015;@grunerbl_smartphone-based_2015;@palmius_detecting_2017]. Other researchers report similar trends in different fields [e.g. @harari_using_2016;@jankowska_framework_2015]. In Figure \@ref(fig:longMeasurementsPerDay) shows that despite the long duration of the log the sparcity it is also evident.

```{r, longMeasurementsPerDay,echo = FALSE,fig.cap="Example of missing data over the entire duration of a secondary log. The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window. For this day there were several periods with no information. These points are filled with red and lie on the x-axis.",out.width = '100%'}

include_graphics("img/missingBoaz5minExample.png")
#change y axis lables to natural numbers, x-axis lables to shorter periods and colour to shapes.

There is no golden standard for dealing with missing data in GPS logs [@barnett_inferring_2016]. Importantly, spatiotemporal data measurements are autocorrelated in both time and space. This means that best practices with other types of data, such as mean imputation, are unsuitable.  For example, imagine an individual who splits almost all her time between work and home. Suppose she spends a small amount of time commuting between the two along a circular path. Using mean imputation to estimate her missing coordinates, we impute her to be at the midpoint between home and work, even though she has never been there. Worryingly, there is little transparency on how researchers deal with missing data [ @jankowska_framework_2015].

Another methodological problem is related to the noise in the measurements that are collected. The accuracy of smartphone location measurements is substantially lower than that of professional GPS location trackers because smartphones often use less accurate sensors. In professional GPS trackers less than 80% of measurements fall within 10 meters of the true location. GPS measures are most inaccurate in dense urban locations and indoors [@schipperijn_dynamic_2014;@duncan_portable_2013]. Unfortunately for researchers, this is where people in the developed world spend most of their time. Figure \@ref(fig:accuracyPlot) shows how accuracy can vary as a function of user behaviour, time and location. Most notably, low accuracy is often (but not always) associated with movement (\@ref(fig:accuracyPlot2)).

```{r, accuracyPlot,echo = FALSE,fig.cap="Measurement accuracy of each logged measurement of a morning journey on February 15th 2017. This includes all measurements from midnight to midday. The red circles denote the accuracy of all logged measurement points (the raw data). The points connected in time are connected by a line. The blue line shows the path without the most inaccurate (accuracy > 400 meters) points filtered out. The red line shows the path with all measurements included. @palmius_detecting_2017 note that in smartphone custom logs inaccurate location values are interspersed between more accurate location values at higher sample rates per hour. Inaccurate measures are often followed by more accurate measures. There are several recurring low-accuracy points, such as the one in the northwest corner, possibly the result of cell-phone tower triangulation.",out.width = '100%'}
include_graphics("img/journeyTillMiddayBoaz.png")
```

@@

Noisy data can lead to inaccurate conclusions if it is not accounted for. Suppose we wish to calculate an individual's movement in a day. A simple approach would be to calculate the sum of the distance between each measurement. But if there is noise, the coordinates will vary even though the individual is not moving. If the measurements are frequent and noisy, we will calculate a lot of movement, even if the individual did not move at all! This issue is also visualised in Figure X. The problem is further complicated because missing data and noisy measurements are related. Methods used by researchers to reduce noise, such as throwing out inaccurate measurements [e.g. @palmius_detecting_2017],  can exacerbate the severity of the missing data problem.

In this paper we will propose PPMI as a method for dealing with missing data and measurement error in secondary location logs. We will compare PPMI to similar solutions in the literature by evaluating the distance between points which were simualted as missing and their imputed  counterparts. Finally we will calculate time spent at home as a function of the imputation method.

@@

How have researchers dealt with missing data in human mobility logs thus far? Unfortunately there is no golden standard in how to deal with this type of missing data. Researchers are generally vague about what practices they follow [@jankowska_framework_2015]. This vagueness is worrisome as it invites solutions which contain significant researcher degrees of freedom [@simmons]. The vaugeness is possibly also due to the fact that most researchers are unfamiliar with possible solutions. Most researchers simply downsample temporally and remove missing observations or use some sort of rule-based common sense imputations (e.g. @palmius_detecting_2017). The only principled approach that we know of that aims to solve the issue of missing data in location logs as they relate to human mobility is that of @barnett_inferring_2016. We will explore the methods of @barnett_inferring_2016  and @palmius_detecting_2017 in detail subsequently, after introducing exploring other spatiotemporal methods.

@@

For secondary logs the main limitation of SSM implementations is that they ignore movement routines.  For instance, humans tend to go to work on weekdays and sleep at night. Because SSMs are based on the Markov property, they cannot incorporate this information. In other words, the estimated location $G(t)$ at timepoint $t$ is often based only upon measurements $D_t$, $D_{t-1}$ and ignores all $D_{t-i}|i\geq2$. Hierarchical structuring and conditioning on a larger context have been suggested as ways to add periodicity to Markovian models. These solutions are often computationally intractable or unfeasible [@sadilek_far_2016]. Moreover, these models often assume time and space invariance (location is not a direct function of time or space). These mathematical assumptions are violated in the case of human movement patterns. For this reason we do not consider existing SSMs to be useful for imputing missing data in this case.

In the wider realm of spatiotemporal statistics there are numerous missing data imputation methods. These often come from climate or geological research and rely on spatiotemporal autocorrelations. For instance, the CUTOFF method estimates missing values by incorporating similar observed temporal information from the value’s nearest spatial neighbors [ @feng_cutoff:_2014 ]. The authors illustrate their example using rainfall data from gauging stations across Australia. Similarly, @zhang_application_2017 use a variety of machine learning methods to impute missing values. The example provided relates to underground water data. Generally these models assume fixed measurement stations (such as rainfall gauging stations). For this reason they cannot be easily applied to missing mobility tracks without significant pre-processing.

On the other hand, a few researchers have explicitly attempted to impute missing data from human mobility patterns [ @palmius_detecting_2017 ;@barnett_inferring_2016; @Wu]. Importantly, none of them worked with secondary logs. Nonetheless we will detail what they did as informative examples. @palmius_detecting_2017 deal with the measuremement inaccuracy of $D$ in custom logs by removing from the data set all unique low-accuracy $a$ data points that had  $\frac{d}{dt}D > 100 \frac{km}{h}$. Subsequently the researchers down sample the data to a sample rate of 12 per hour using a median filter. Moreover, @palmius_detecting_2017 explain:

@@

@barnett_inferring_2016 follow a different approach which is, to the best of our knowledge, the only pricipled approach to dealing with missing data in human mobility data. @barnett_inferring_2016 work with custom logs where location is measured for 2 minutes and subsequently not measured for 10 minutes. In the words of the authors, @barnett_inferring_2016 handle missing data by first converting data to mobility traces, which are defined as a sequence of flights and pauses. Flights are segments of linear movements and pauses corresponding to periods of time where a person does not move. Subsequently, the authors impute missing data by:

@@

On the other hand, @wu use what they call a Spatial Temporal Semantic Neural Network (STS-NN) to predict future human movement. While the authors are concerned with prediction and not imputation, they devised a method called the Spatial Temporal Semantic (STS) algorithm which converts raw measurements to machine learning friendly discrete bins. Working with high-frequency measurements, @wu's method downsamples the raw data temporally and map-matches the resulting bins to discrete points along pre-established geographical features such as roads and highways. This minimises measurement error and paves the way for applying machine learning methods to human mobility problems.

@@

GPS measurements provide us with coordinates on the surface of the earth, which is ellipsoid shaped. Projecting three dimensional measurements in $\mathbb{R}^3$ onto a two dimensional plane in $\mathbb{R}^2$ results in distortion. For clarity, when we use the term distance we refer to the geodesic distances on an ellipsoid using the WGS84 ellipsoid parameters.

Subsequently let us simplify by assuming that a persons location is on two-dimensional Euclidian plane. Let a person's true location on this two-dimensional plane be $G(t) = [G_x(t) G_y(t)]$ where $G_x(t)$ and $G_y(t)$ denote the location of the individual at time $t$ on the x-axis and y-axis respectively. For simplicity, we can assume that the x-axis is the longitude and the y-axis is the lattitude. Moreover, let $D \in \mathbb{R}^2$ be the recorded data containing the lattitude and longitude. In addition, let $a$ denote the estimated accuracy of the recorded data. Furthermore, $G(t)$, $D$ and $a$ are indexed by time labled by the countable set $t = t_1 < ... < t_{n+1}$. The measure of accuracy $a_t$ is given in meters such that it represents the radius of a 67% confidence circle. If $D_t = \emptyset$ it is considered *missing* and it is not missing otherwise.

@@

### Personalised Map Matching Imputation

Our algorithm is designed to leverage the periodic nature of human movement along with the long span of secondary to deal with measurement sparcity and inaccuracy.

@@

First, following @barnett_inferring_2016 we categorise all timepoints $t$ as either belonging to the set $P$ (pause) or set $F$ (flight). Conceptually pauses can be understood as periods of time where an individual spends singificant amount of continous time without moving. Flights are the times where the individual is moving. Let $t_a$ be a pause of length $n$.
$$t_a = t_i < ... < t_{i+n}$$
Let $t_b$ be a pause of length $m$ such that there is no temporal overlap between $t_a$ and $t_b$:
$$t_b = t_j= < ... < t_{j+m}|t_{i+n}< t_j$$
Then it follows that between the two pauses there must be a flight indexed by $t_x$ of length $j-i+n$.
$$t_x =  t_{i+n} < ... < t_{j} |t_x \in F$$
We define pause locations $G(t_a), G(t_b) | t_a, t_b \in P$ as locations where an individual spends an extended amount of time in the same space (e.g. school, home, work, train station, barber shop, bar, gym). Importantly, our model assumes period and cyclic human movement such that there are many pauses $t_{a1},t_{a2},...,t_{an}$ such that $G(t_{a1}) = G(t_{a2}) = ... = G(t_{an})$. Moreover,it is possible for $G(t_a) = G(t_b)$ such that $t_a \ne t_b$. For example, if the individual leaves home for a run and returns home without stopping anywhere else.

Let us define as $Flight^x_{ab}$ the set of all points belonging to a flight between $G(t_a)$ and $G(t_b)$ at timepoint $t_x$.
$$Flight^x_{ab}= G(t_x)|t_x \in F =  \{G(t_{i+n}),...,G(t_j)\}$$
Again, there are many flights $t_{x1},t_{x2},...,t_{xn}$ such that $Flight^{x1}_{ab} = Flight^{x2}_{ab}  = ... = Flight^{xn}_{ab}$. Then, we can define as $Path_{ab}$ the set of all flights between $G(t_a)$ and $G(t_b)$ at all timepoints. For simplicity, we assume that $Path_{ab}=Path_{ba}$.

@@

### Personalised Map Matching Imputation algorithm

@@

1. Map building: Extract from measurements $D$ all pause locations and path clusters to create a personalised map.
2. Binning: Assign each measurement $D$ to a unique discrete location. 
3. Imputing: Use a classification method to predict missing measurements based on the time as well as the distance in time to the next and previous measurements.

#### Map building

Following @wu's spatial-temporal-semantic (STS) feature extraction algorithm our aim is to transform pause and path locations into machine learning friendly discrete location sequences. There are multiple ways of extracting such measurement clusters in the literature, such as Spatio-Temporal Density-Based Spatial Clustering of Applications with Noise (ST-DBSCAN) and sequence oriented clustering (SOC) [@xiang;]. We will focus on two methods which explicitly with mobility patterns from unevenly sampled smartphone logs [ @palmius_detecting_2017 ;@barnett_inferring_2016]. Both of these methods pre-process the data and subsequently use two steps to extract pause locations: first they extract pauses and their corresponding locations, then they cluster pause locations based on spatial proximity. For simplicity this implementation of PMMI uses a stricter version of @barnett_inferring_2016's approach to extract pauses. However, there is no reason why more sophisticated methods cannot be used.

@@

This set of points were then hierarchically clustered using a distance matrix, such that all points within $d$ meters of each other were clustered into a pause location. Each pause location is a cluster.

For all remaining points we assume that they belong to paths. In this implementation of PMMI we use the following algorithm to estimate paths:

>>>>

---
title             : "Handling missing data in smartphone location logs"
shorttitle        : "Missing Data in Smartphone Location Logs"

author: 
  - name          : "Boaz Sobrado"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : ""
    email         : "boaz@boazsobrado.com"

affiliation:
  - id            : "1"
    institution   : "Utrecht University"

author_note: |
  Department of Methodology & Statistics

  Submitted as a research report conforming to APA manuscript guidelines (6th edition).

abstract: |
  Personal mobility, or how people move in their environment, is associated with a vast range of behavioural traits and outcomes, such as socioeconomic status, personality and mental health. The widespread adoption of location-sensor equipped smartphones has generated a wealth of objective personal mobility data. Nonetheless, smartphone collected personal mobility data has remained underutilised in behavioural research, partly due to the practical difficulties associated with obtaining the data and partly because of the methodological complexity associated with analysing it. Recent changes in European regulation have made it easier for researchers to obtain this data, but the methodological difficulties remain. The difficulty lies in that smartphone location data is irregularly sampled, sparse and often inaccurate.  This results in a high proportion of missing data and significant noise. In this paper we present a method called Personal Map Matched Imputation (PPMI) to deal with missing data and noise in smartphone location logs. The main innovation of PPMI is that it creates a personalised spatial map for each individual based on all the available data. In doing so PPMI leverages the regularity of human mobility in order to smoothen noisy measurments and impute missing data values. By simulating missing periods in real data we find that a simple implementation of PPMI performs as well as existing methods for short (5 minute) missing intervals and substantially better for longer (1 day) missing intervals. When imputing a subset of real  missing data where travel logs are available as a reference points, we find that PPMI performs substantially better than existing models.

keywords          : "Missing Data, Measurement Bias, GPS, Human Mobility"
wordcount         : "2297"

bibliography      : 
  - thesisRef.bib
  - r-references.bib

figsintext        : no
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : yes
mask              : no

class             : "man"
output            : papaja::apa6_pdf

---

```{r load_packages, include = FALSE}
library("papaja")
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(ggthemes)
library(ggplot2)
library(knitr)
library(kableExtra)

@@

Why is human mobility important? How people move about in their environment affects a wide range of outcomes, such as health, income and social capital [@goodchild_toward_2010]. Therefore it is unsurprising that social scientists in numerous fields and even policymakers are interested in human mobility measures. The most widely administered personality questionnaires ask individuals to what extent they agree with statements such as "I love large parties", "I prefer going to the movies to watching videos at home" and "I love to travel to places that I have never been before" [@ipip]. In economic research, the postal code of an individual's home address is often used as a proxy for socioeconomic status [e.g. @villanueva_association_2013]. Moreover, economists study geographic labour mobility extensively [e.g. @tatsiramos]. In fact, the European Commission considers labour mobility important enough to warrant its own directorate [@europeanComission]. The extensive use of measures like these across different domains strongly suggests that mobility metrics are linked to real world outcomes. Perhaps it is unsurprising that behavioural researchers have found that mobility measures can be used to predict academic performance [@wang_smartgpa:_2015], the incidence of obesity [@zenk_how_2009] or even the onset of a depressive episode in bipolar depression patients [@palmius_detecting_2017]. Indeed, there is an argument to be made that perhaps psychologists have not been studying mobility enough. When studying behavioural differences within individuals behavioural scientists have often neglected the fact that individuals vary not only over time but also over space. To fully understand behaviour we must understand how behaviour can vary across environments.

Despite the importance of mobility measures, the majority of these measures are obtained through the use of questionnaires (such as the IPIP) or specifically through pen-and-paper travel diaries. Questionnaires and travel diaries are have well-known methodological flaws: they are burdensome to collect and rely on accurate self-reporting. The burdensomeness stems from the fact that participants must be explicitly asked to report on their movement patterns at frequent intervals due to forgetfulness. This makes the data expensive to collect and limits the extent of data collection in practice. In addition, the frequent reporting duties of the participant may bias the participants behaviour. Participant forgetfulness also limits the accuracy of self reported measures. There is clear evidence that participants are systematically biased when self-reporting mobility measures. For instance, participants under-report the frequency of short trips [@wolf_impact_2003] and underestimate the duration of regular commutes [@delclos-alio_keeping_2017]. These obstacles can be overcome by using objective data on human mobility.

Social scientists now have the unprecedented opportunity to easily obtain objective data on human mobility from smartphones. Before smartphones the only way to collect objective data on human mobility involved giving participants an expensive professional-grade location sensor and convincing them to take it with themselves at all times. @barnett_inferring_2016 points out that introducing a new device to the participant's life may bias their behaviour. Moreover, collecting data in such a way is costly, places a high burden on participants and therefore the logs do not span a long time [@barnett_inferring_2016. Today millions of individuals carry smartphones with themselves every day and do not need to be encouraged to do so by researchers. Smartphones are equipped with a range of sensors that can be used to track the location of the device at all times. These smartphones can collect and store hundreds of location measurements a day. For instance,  Google Location History contains movement information on millions of users, often spanning years [@location_history_timeline_2017]. Moreover, recent changes in EU regulations with regard to consumer data-portability rights ensure that a willing participant should be able to easily share this information with researchers at no cost to the either the participant or the researchers [@commission_protecting_2017]. Taken together, this means that researchers now have at their disposal the ability to easily access the objective human mobility data of millions of individuals spanning several years at little cost and without a significant burden of the participants.

@@

Smartphone location measures are obtained primarily (but not exclusively) by Global Positioning System (GPS) measurements. A GPS sensor uses the distance between a device and several satellites to determine the location of the device. Although using a GPS sensor is the most accurate way to establish location on a smartphone, the GPS sensor is also the most energy consuming sensor on most smartphones [@lamarca_place_2005; @chen_practical_2006]. In order to avoid battery depletion and to overcome computational constraints smartphones also use less-accurate heuristics such as WiFi access points and cellphone tower triangulation. Smartphone location logs contain measurements from all of these sources, usually in the form of time-stamped latitude, longitude and accuracy values. The accuracy $a$ of any given measurement is given in meters such that it represents the radius of a 67% confidence circle [@location_history_timeline_2017]. In other words, the true location of a device should be within the radius $a$ of the measurement 67% of the time.

@@

In order to work with secondary logs, researchers need to be able to handle the data sparsity that leads to missing data. Missing data is a pervasive issue in secondary logs as it can arise due to several reasons. Technical reasons include signal loss, battery failure and device failure. Behavioural reasons include leaving the phone at home or switching the device off. As a result, secondary logs often contain wide temporal gaps with no measurements. For instance, several research groups studying mental health report missing data rates between 30% to 50% [@saeb_mobile_2015;@grunerbl_smartphone-based_2015;@palmius_detecting_2017]. Other researchers report similar trends in different fields [e.g. @harari_using_2016;@jankowska_framework_2015]. In Figure \@ref(fig:longMeasurementsPerDay) shows that despite the long duration of the log the sparsity it is also evident.

```{r, longMeasurementsPerDay,echo = FALSE,fig.cap="Example of missing data over the entire duration of a secondary log. The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window. For this day there were several periods with no information. These points lie on the x-axis.",out.width = '100%'}

include_graphics("img/missin5minExample2018.png")
```

There is no golden standard for dealing with missing data in GPS logs [@barnett_inferring_2016]. Importantly, spatiotemporal data measurements are auto-correlated in both time and space. This means that best practices with other types of data, such as mean imputation, are unsuitable.  For example, imagine an individual who splits almost all her time between work and home. Suppose she spends a small amount of time commuting between the two along a circular path. Using mean imputation to estimate her missing coordinates, we impute her to be at the midpoint between home and work, even though she has never been there. Worryingly, there is little transparency on how researchers deal with missing data [ @jankowska_framework_2015].

Another methodological problem is related to the noise in the measurements that are collected. The accuracy of smartphone location measurements is substantially lower than that of professional GPS location trackers because smartphones often use less accurate sensors. In professional GPS trackers less than 80% of measurements fall within 10 meters of the true location. GPS measures are most inaccurate in dense urban locations and indoors [@schipperijn_dynamic_2014;@duncan_portable_2013]. Unfortunately for researchers, this is where people in the developed world spend most of their time. Figure \@ref(fig:accuracyPlot) shows how accuracy can vary as a function of user behaviour, time and location. Most notably, low accuracy is often (but not always) associated with movement (see Figure \@ref(fig:accuracyPlot2)).

```{r, accuracyPlot,echo = FALSE,fig.cap="Measurement accuracy of each logged measurement of a morning journey on February 15th 2017. This includes all measurements from midnight to midday. The red circles denote the accuracy of all logged measurement points (the raw data). The points connected in time are connected by a line. The blue line shows the path without the most inaccurate (accuracy > 400 meters) points filtered out. The red line shows the path with all measurements included. In smartphone logs inaccurate location values are interspersed between more accurate location values at higher sample rates per hour. Inaccurate measures are often followed by more accurate measures. There are several recurring low-accuracy points, such as the one in the northwest corner, possibly the result of cellphone tower triangulation.",out.width = '100%'}
include_graphics("img/journeyTillMiddayBoaz.png")
```

@@

Noisy data can lead to inaccurate conclusions if it is not accounted for. Suppose we wish to calculate an individual's movement in a day. A simple approach would be to calculate the sum of the distance between each measurement. But if there is noise, the coordinates will vary even though the individual is not moving. If the measurements are frequent and noisy, we will calculate a lot of movement, even if the individual did not move at all! This issue is also visualised in Figure \@ref(fig:palmiusVme1). The problem is further complicated because missing data and noisy measurements are related. Methods used by researchers to reduce noise, such as throwing out inaccurate measurements [e.g. @palmius_detecting_2017],  can exacerbate the severity of the missing data problem.

In this paper we will propose PPMI as a method for dealing with missing data and measurement error in secondary location logs. We will compare PPMI to similar solutions in the literature by evaluating the distance between points which were simulated as missing and their imputed  counterparts. Finally we will calculate time spent at home as a function of the imputation method.

@@

How have researchers dealt with missing data in human mobility logs thus far? Unfortunately there is no golden standard in how to deal with this type of missing data. Researchers are generally vague about what practices they follow [@jankowska_framework_2015]. This vagueness is worrisome as it invites solutions which contain significant researcher degrees of freedom [@simmons_false-positive_2011]. The vagueness is possibly also due to the fact that most researchers are unfamiliar with possible solutions. Most researchers simply down-sample temporally and remove missing observations or use some sort of rule-based common sense imputations (e.g. @palmius_detecting_2017). The only principled approach that we know of that aims to solve the issue of missing data in location logs as they relate to human mobility is that of @barnett_inferring_2016. We will explore the methods of @barnett_inferring_2016  and @palmius_detecting_2017 in detail subsequently, after introducing exploring other spatiotemporal methods.

@@

For secondary logs the main limitation of SSM implementations is that they ignore movement routines.  For instance, humans tend to go to work on weekdays and sleep at night. Because SSMs are based on the Markov property, they cannot incorporate this information. In other words, the estimated location $G(t)$ at time-point $t$ is often based only upon measurements $D_t$, $D_{t-1}$ and ignores all $D_{t-i}|i\geq2$. Hierarchical structuring and conditioning on a larger context have been suggested as ways to add periodicity to Markovian models. These solutions are often computationally intractable or unfeasible [@sadilek_far_2016]. Moreover, these models often assume time and space invariance (location is not a direct function of time or space). These mathematical assumptions are violated in the case of human movement patterns. For this reason we do not consider existing SSMs to be useful for imputing missing data in this case.

In the wider realm of spatiotemporal statistics there are numerous missing data imputation methods. These often come from climate or geological research and rely on spatiotemporal auto-correlations. For instance, the CUTOFF method estimates missing values by incorporating similar observed temporal information from the value’s nearest spatial neighbors [ @feng_cutoff:_2014 ]. The authors illustrate their example using rainfall data from gauging stations across Australia. Similarly, @zhang_application_2017 use a variety of machine learning methods to impute missing values. The example provided relates to underground water data. Generally these models assume fixed measurement stations (such as rainfall gauging stations). For this reason they cannot be easily applied to missing mobility tracks without significant pre-processing.

On the other hand, a few researchers have explicitly attempted to impute missing data from human mobility patterns [ @palmius_detecting_2017 ;@barnett_inferring_2016; @Wu]. Importantly, none of them worked with secondary logs. Nonetheless we will detail what they did as informative examples. @palmius_detecting_2017 deal with the measurement inaccuracy of $D$ in custom logs by removing from the data set all unique low-accuracy $a$ data points that had  $\frac{d}{dt}D > 100 \frac{km}{h}$. Subsequently the researchers down sample the data to a sample rate of 12 per hour using a median filter. Moreover, @palmius_detecting_2017 explain:

@@

@barnett_inferring_2016 follow a different approach which is, to the best of our knowledge, the only principled approach to dealing with missing data in human mobility data. @barnett_inferring_2016 work with custom logs where location is measured for 2 minutes and subsequently not measured for 10 minutes. In the words of the authors, @barnett_inferring_2016 handle missing data by first converting data to mobility traces, which are defined as a sequence of flights and pauses. Flights are segments of linear movements and pauses corresponding to periods of time where a person does not move. Subsequently, the authors impute missing data by:

@@

On the other hand, @wu use what they call a Spatial Temporal Semantic Neural Network (STS-NN) to predict future human movement. While the authors are concerned with prediction and not imputation, they devised a method called the Spatial Temporal Semantic (STS) algorithm which converts raw measurements to machine learning friendly discrete bins. Working with high-frequency measurements, @wu's method down-samples the raw data temporally and map-matches the resulting bins to discrete points along pre-established geographical features such as roads and highways. This minimises measurement error and paves the way for applying machine learning methods to human mobility problems.

@@

Location measurements, such as those produced by GPS sensors, provide us with coordinates (latitude and longitude) on the surface of the earth, which is ellipsoid shaped. Projecting three dimensional measurements in $\mathbb{R}^3$ onto a two dimensional plane in $\mathbb{R}^2$ results in distortion. For clarity, when we use the term distance we refer to the geodesic distances on an ellipsoid using the WGS84 ellipsoid parameters.

Subsequently let us simplify by assuming that a persons location is on two-dimensional Euclidean plane. Let a person's true location on this two-dimensional plane be $G(t) = [G_x(t) G_y(t)]$ where $G_x(t)$ and $G_y(t)$ denote the location of the individual at time $t$ on the x-axis and y-axis respectively. For simplicity, we can assume that the x-axis is the longitude and the y-axis is the latitude. Moreover, let $D \in \mathbb{R}^2$ be the recorded data containing the latitude and longitude. In addition, let $a$ denote the estimated accuracy of the recorded data. Furthermore, $G(t)$, $D$ and $a$ are indexed by time labeled by the countable set $t = t_1 < ... < t_{n+1}$. The measure of accuracy $a_t$ is given in meters such that it represents the radius of a 67% confidence circle. If $D_t = \emptyset$ it is considered *missing* and it is not missing otherwise.

@@

## Personalised Map Matching Imputation

Our algorithm is designed to leverage the periodic nature of human movement along with the long span of secondary to deal with measurement sparsity and inaccuracy.

@@

First, following @barnett_inferring_2016 we categorise all time-points $t$ as either belonging to the set $P$ (pause) or set $F$ (flight). Conceptually pauses can be understood as periods of time where an individual spends significant amount of continuous time without moving. Flights are the times where the individual is moving. Let $t_a$ be a pause of length $n$.
$$t_a = t_i < ... < t_{i+n}$$
Let $t_b$ be a pause of length $m$ such that there is no temporal overlap between $t_a$ and $t_b$:
$$t_b = t_j= < ... < t_{j+m}|t_{i+n}< t_j$$
Then it follows that between the two pauses there must be a flight indexed by $t_x$ of length $j-i+n$.
$$t_x =  t_{i+n} < ... < t_{j} |t_x \in F$$
We define pause locations $G(t_a), G(t_b) | t_a, t_b \in P$ as locations where an individual spends an extended amount of time in the same space (e.g. school, home, work, train station, barber shop, bar, gym). Importantly, our model assumes period and cyclic human movement such that there are many pauses $t_{a1},t_{a2},...,t_{an}$ such that $G(t_{a1}) = G(t_{a2}) = ... = G(t_{an})$. Moreover,it is possible for $G(t_a) = G(t_b)$ such that $t_a \ne t_b$. For example, if the individual leaves home for a run and returns home without stopping anywhere else.

Let us define as $Flight^x_{ab}$ the set of all points belonging to a flight between $G(t_a)$ and $G(t_b)$ at time-point $t_x$.
$$Flight^x_{ab}= G(t_x)|t_x \in F =  \{G(t_{i+n}),...,G(t_j)\}$$
Again, there are many flights $t_{x1},t_{x2},...,t_{xn}$ such that $Flight^{x1}_{ab} = Flight^{x2}_{ab}  = ... = Flight^{xn}_{ab}$. Then, we can define as $Path_{ab}$ the set of all flights between $G(t_a)$ and $G(t_b)$ at all time-points. For simplicity, we assume that $Path_{ab}=Path_{ba}$.

@@

## Personalised Map Matching Imputation algorithm

@@

1. *Map building*: Extract from measurements $D$ all pause location bins and path location bins to create a personalised map.
2. *Binning*: Assign each measurement $D$ to a unique discrete location bin. 
3. *Imputing*: Use a classification method to predict missing measurements based on all the available information.

### Map building

Following @wu's spatial-temporal-semantic (STS) feature extraction algorithm our aim is to transform pause and path locations into machine learning friendly discrete location sequences. There are multiple ways of extracting such measurement clusters in the literature, such as Spatio-Temporal Density-Based Spatial Clustering of Applications with Noise (ST-DBSCAN) and sequence oriented clustering (SOC) [@_st-dbscan:_2007;]. We will focus on two methods which explicitly with mobility patterns from unevenly sampled smartphone logs [ @palmius_detecting_2017 ;@barnett_inferring_2016]. Both of these methods pre-process the data and subsequently use two steps to extract pause locations: first they extract pauses and their corresponding locations, then they cluster pause locations based on spatial proximity. This implementation of PMMI uses a stricter version of @barnett_inferring_2016's approach to extract pauses.

@@

This set of points were then hierarchically clustered using a distance matrix, such that all points within $d$ meters of each other were clustered into a pause location. Each pause location is a bin.

For all remaining measurements we assume that they belong to paths. In this implementation of PMMI we use the following algorithm to estimate paths:
